Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-08-18 prop_letter.tex
prop_letter.tex \section{Kochen-Specker}  Change N to M...
lots of small changes

The paper is organized as follows.  In Sec.\ \ref{sec:fuzzy} we begin
by introducing the foundations of quantum probability space and fuzzy
measurement.  We then move on to introduce the concept of quantum
interval valued probability in Sec.\ \ref{sec:Interval-Uncertainty}.
The QIVPM will provide the mathematical framework we need to quantify the
impact of finite precision measurements on quantum mechanics.  For
instance, in Sec.\ \ref{sec:Kochen-Specker}, we quantify the domain of
validity of a contextual measurement, thus addressing the conditions
under which the Kochen-Specker theorem applies.  This provides a way
not only of resolving the Meyer-Mermin debate, but also revealing a
precise transition to non-contextuality;  these are the results of our
Thm~\ref{cor:Kochen-Specker-IVPM}.  In Sec.\ \ref{sec:Gleason}, we
study Gleason's theorem in the context of imprecise measurements,
concluding that QIVPM provides a framework with quantitative bounds in
which Gleason's theorem, while formally invalid in a universe with
bounded resources, holds asymptotically in a mathematically precise
way.  Finally, we present our conclusions in Sec.~\ref{sec:Conclusion}.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$,
the identity projector~$\mathbb{1}$, projectors of the form
$\proj{\phi}$ where $\ket{\phi}$ is a pure quantum state (an element
of a Hilbert space), sums of \emph{orthogonal} projectors $P_0$
and~$P_1$ with $P_0P_1=\mathbb{0}$, and products of \emph{commuting}
projectors $P_0$ and~$P_1$ with $P_0P_1=P_1P_0$. In a well-defined
quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$
is the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given an observable~$\mathbf{O}$ with spectral decomposition
$\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ one can define the 
expectation value of that observable as:
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state~$\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

@@

We will start by reviewing IVPMs in the classical setting and then
propose our quantum generalization. In the classical setting, there
are several proposals for ``imprecise
probabilities''~\citep{Dempster1967,Shafer1976,GilboaSchmeidler1994,Marinacci1999,Weichselberger2000,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
Although these proposals differ in some details, they all share the
fact that the probability $\bar{\mu}(E)$ of an event~$E$ is generalized from
a single \emph{real number} to an \emph{interval}~$[\ell,r]$
where~$\ell$ intuitively corresponds to the strength of evidence for
the event~$E$ and~$1-r$ corresponds to the strength of evidence
against the same event. The collection of intervals used in the
definition of a probability measure must satisfy certain properties
that we explain next.

First, for each interval $[\ell,r]$ we have the natural constraint
$0 \leq \ell \leq r \leq 1$ that guarantees that every element of the
interval can be interpreted as a conventional probability. We also
include $\imposs=[0,0]$ and $\necess=[1,1]$ as limiting intervals to
refer to the probability of impossible events (e.g., the null event)
and events that are certain (e.g., the event covering the entire
sample space).  For each interval $[\ell,r]$, we also need the
interval $[1-r,1-\ell]$ so that if one interval refers to the
probability of an event, the other refers to the probability of the
event's complement.  For example, if we discover as a result of an
experiment that $\bar{\mu}(E) = [0.2,0.3]$ for some event~$E$, we may
conclude that $\bar{\mu}(\overline{E}) = [0.7,0.8]$ for the complement event
$\overline{E}$. In addition to these simple conditions, there are some
subtle conditions on how intervals are combined, which we discuss
next.

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\bar{\mu}(E_1)=[\ell_1,r_1]$ and $\bar{\mu}(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\bar{\mu}(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\bar{\mu}(E_1)=[0.1,0.2]$ and
$\bar{\mu}(E_2)=[0.3,0.4]$ we get $\bar{\mu}(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\bar{\mu}(E)=[0.2,0.3]$ and hence
$\bar{\mu}(\overline{E})=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the previous naïve way of combining the intervals would
give $\bar{\mu}(E\cup\overline{E})=[0.9,1.1]$ which is not a valid
probability interval. Moreover the event $E\cup\overline{E}$ is the
identity event covering the entire space; its probability interval
should be $[1,1]$ which is sharper than $[0.9,1.1]$. The problem is
that the two intervals are correlated: there is more information in
the combined event than in each event separately. In our example, even
though the true probability of {$\bar{\mu}(E)$}, $\mu(E)$, can be anywhere in the range
$[0.2,0.3]$ and the true probability of {$\bar{\mu}(\overline{E})$}, $\mu(\overline{E})$, can be in
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu(\overline{E})\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint regarding the probability of
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is
$\bar{\mu}(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]$ which, in
combination with the requirement for the identity event would force
$\bar{\mu}(E \cup \overline{E})$ to always be $[1,1]$ irrespective of the
particular assignments of $\bar{\mu}(E)$ or~$\bar{\mu}(\overline{E})$.

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
general, not necessarily disjoint, events $E_1$ and $E_2$, we have:
\begin{equation}
\bar{\mu}(E_1\cup E_2) + \bar{\mu}(E_1\cap E_2) \subseteq \bar{\mu}(E_1) + \bar{\mu}(E_2)
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as
\emph{convexity}~\cite{Shapley1971,GilboaSchmeidler1994,Marinacci1999,Grabisch2016},
reduces to the previously
motivated condition when the events are disjoint, i.e., when
$\bar{\mu}(E_1\cap E_2) = 0$.

@@

\begin{definition}[Expectation Value of Observables over QIVPMs] Let
  $\mathscr{I}$ be a set of intervals; $\Hilb$ a Hilbert space of
  dimension~$d\ge3$ with event space $\events$; and $\mathbf{O}$ an
  observable with spectral decomposition
  $\Sigma_{i=1}^N \lambda_iP_i$. Let $\events'$ be the minimal
  subspace of events containing all the projectors $P_i$, then we define:
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

@@

Our generalization of quantum probability measures to QIVPMs will
allow us to strengthen one of the fundamental theorems in of quantum
information: the Kochen-Specker
theorem~\cite{PeresRon1988,peres1995quantum}. Our stronger version of
the theorem will suggest a resolution to the debate initiated by Meyer
and Mermin on the relevance of the Kochen-Specker to experimental, and
hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
Specifically, the original Kochen-Specker theorem is formulated
under the assumption that all observables of a quantum mechanical
system have \emph{definite values at all
  times}~\cite{Held2016}. Our interval-valued framework will allow
us to state, and prove, a stronger version of the theorem that holds
even if the observables have values that are only definite up to some
specific parameter that quantifies experimental uncertainty. Put
differently, our stronger statement and theorem quantifies Mermin's
qualitative argument:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

\paragraph*{Expectation Values of Fuzzy Observables.}  Our following
development will crucially rely on analyzing the expectation values of
families of observables. We introduce the necessary definitions before
stating and proving our variants of the Kochen-Specker theorem.

The expectation value is intuitively between two possible outcomes,
which themselves lie between the minimum and maximum bounds of the
given probability intervals. (This definition is consistent with the
classical Choquet integral on subspaces of commuting events by
adopting the classical
results~\cite{GilboaSchmeidler1994,Grabisch2016}.)  If the QIVPM is
0-deterministic, this interval collapses to a singleton
$\left\{ \lambda\right\} $, where $\lambda$ is an eigenvalue
of~$\mathbf{O}$. If we define the multiplication of two singleton
sets~$\left\{ \lambda_{1}\right\} $ and $\left\{ \lambda_{2}\right\} $
to be $\left\{ \lambda_{1}\lambda_{2}\right\} $, then we have
\begin{equation}
\prod_{j=1}^{N}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{N}\mathbf{O}_{j}}_{\bar{\mu}}\label{eq:Kochen-Specker-Rules}
\end{equation}
for any set of commuting observables~$\left\{ \mathbf{O}_{j}\right\}
_{j=1}^{N}$.

@@

>>>>

The paper is organized as follows.  In Sec.~\ref{sec:fuzzy} we begin
by introducing the foundations of quantum probability space and fuzzy
measurement.  We then move on to introduce the concept of quantum
interval valued probability in Sec.~\ref{sec:Interval-Uncertainty}.
The QIVPM will provide the mathematical framework we need to quantify the
impact of finite precision measurements on quantum mechanics.  For
instance, in Sec.~\ref{sec:Kochen-Specker}, we quantify the domain of
validity of a contextual measurement, thus addressing the conditions
under which the Kochen-Specker theorem applies.  This provides a way
not only of resolving the Meyer-Mermin debate, but also revealing a
precise transition to non-contextuality;  these are the results of our
Thm~\ref{cor:Kochen-Specker-IVPM}.  In Sec.~\ref{sec:Gleason}, we
study Gleason's theorem in the context of imprecise measurements,
concluding that QIVPM provides a framework with quantitative bounds in
which Gleason's theorem, while formally invalid in a universe with
bounded resources, holds asymptotically in a mathematically precise
way.  Finally, we present our conclusions in Sec.~\ref{sec:Conclusion}.

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$, the
identity projector~$\mathbb{1}$, projectors of the form $\proj{\phi}$
where $\ket{\phi}$ is a pure quantum state (an element of a Hilbert
space), sums of \emph{orthogonal} projectors $P_0$ and~$P_1$ with
$P_0P_1=\mathbb{0}$, and products of \emph{commuting} projectors $P_0$
and~$P_1$ with $P_0P_1=P_1P_0$. In a quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$ is
the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given a probability for every projector, we can associate an
expectation value with an observable~$\mathbf{O}$ with spectral
decomposition $\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ as:
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state~$\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\vert\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

@@

We will start by reviewing IVPMs in the classical setting and then
propose our quantum generalization. In the classical setting, there
are several proposals for ``imprecise
probabilities''~\citep{Dempster1967,Shafer1976,GilboaSchmeidler1994,Marinacci1999,Weichselberger2000,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
Although these proposals differ in some details, they all share the
fact that the probability $\mu(E)$ of an event~$E$ is generalized from
a single \emph{real number} to an \emph{interval}~$[\ell,r]$
where~$\ell$ intuitively corresponds to the strength of evidence for
the event~$E$ and~$1-r$ corresponds to the strength of evidence
against the same event. The collection of intervals used in the
definition of a probability measure must satisfy certain properties
that we explain next.

First, for each interval $[\ell,r]$ we have the natural constraint
$0 \leq \ell \leq r \leq 1$ that guarantees that every element of the
interval can be interpreted as a conventional probability. We also
include $\imposs=[0,0]$ and $\necess=[1,1]$ as limiting intervals to
refer to the probability of impossible events (e.g., the null event)
and events that are certain (e.g., the event covering the entire
sample space).  For each interval $[\ell,r]$, we also need the
interval $[1-r,1-\ell]$ so that if one interval refers to the
probability of an event $E$, the other refers to the probability of
the event's complement $\overline{E}$.  For example, if we discover as
a result of an experiment that $\mu(E) = [0.2,0.3]$ for some
event~$E$, we may conclude that $\mu(\overline{E}) = [0.7,0.8]$ for
the complement event $\overline{E}$. In addition to these simple
conditions, there are some subtle conditions on how intervals are
combined, which we discuss next.

Let $E_1$ and $E_2$ be two disjoint events with probabilities
$\mu(E_1)=[\ell_1,r_1]$ and $\mu(E_2)=[\ell_2,r_2]$. A first attempt
at calculating the probability of the combined event that
\emph{either} $E_1$ or $E_2$ occurs might be
$\mu(E_1\cup E_2) = [\ell_1+\ell_2,r_1+r_2]$. In some cases, this is
indeed a sensible definition. For example, if $\mu(E_1)=[0.1,0.2]$ and
$\mu(E_2)=[0.3,0.4]$ we get $\mu(E_1\cup E_2) = [0.4,0.6]$. But
consider an event $E$ such that $\mu(E)=[0.2,0.3]$ and hence
$\mu(\overline{E})=[0.7,0.8]$. The two events $E$ and $\overline{E}$
are disjoint; the previous naïve way of combining the intervals would
give $\mu(E\cup\overline{E})=[0.9,1.1]$ which is not a valid
probability interval. Moreover the event $E\cup\overline{E}$ is the
identity event covering the entire space; its probability interval
should be $[1,1]$ which is sharper than $[0.9,1.1]$. The problem is
that the two intervals are correlated: there is more information in
the combined event than in each event separately. In our example, even
though the ``true'' probability of $E$ can be anywhere in the range
$[0.2,0.3]$ and the ``true'' probability of $\overline{E}$ can be in
anywhere in the range $[0.7,0.8]$, the values are not independent. Any
value of $\mu(E) \leq 0.25$ will force $\mu(\overline{E})\geq
0.75$. To account for such subtleties, the axioms of interval-valued
probability do not use a strict equality for the combination of
disjoint events. The correct constraint regarding the probability of
$E_1\cup E_2$ when $E_1$ and $E_2$ are disjoint is
$\mu(E_1\cup E_2) \subseteq [\ell_1+\ell_2,r_1+r_2]$ which, in
combination with the requirement for the identity event would force
$\mu(E \cup \overline{E})$ to always be $[1,1]$ irrespective of the
particular assignments of $\mu(E)$ or~$\mu(\overline{E})$.

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
general, not necessarily disjoint, events $E_1$ and $E_2$, we have:
\begin{equation}
\mu(E_1\cup E_2) + \mu(E_1\cap E_2) \subseteq \mu(E_1) + \mu(E_2)
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as
\emph{convexity}~\cite{Shapley1971,GilboaSchmeidler1994,Marinacci1999,Grabisch2016},
reduces to the previously
motivated condition when the events are disjoint, i.e., when
$\mu(E_1\cap E_2) = 0$.

@@

\begin{definition}[Expectation Value of Observables over QIVPMs] Let
  $\mathscr{I}$ be a set of intervals; $\Hilb$ a Hilbert space of
  dimension~$d\ge3$ with event space~$\events$; and $\mathbf{O}$ an
  observable with spectral decomposition
  $\Sigma_{i=1}^N \lambda_iP_i$. Let $\events'$ be the minimal
  subspace of events containing all the projectors $P_i$, then we define:
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

@@

Our generalization of quantum probability measures to QIVPMs allows us
to strengthen one of the fundamental theorems in of quantum
information: the Kochen-Specker
theorem~\cite{PeresRon1988,peres1995quantum}. Our stronger version of
the theorem will suggest a resolution to the debate initiated by Meyer
and Mermin on the relevance of the Kochen-Specker to experimental, and
hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
Specifically, the original Kochen-Specker theorem is formulated under
the assumption that all observables of a quantum mechanical system
have \emph{definite values at all times}~\cite{Held2016}. Our
interval-valued framework will allow us to state, and prove, a
stronger version of the theorem that holds even if the observables
have values that are only definite up to some specific parameter that
quantifies experimental uncertainty. Put differently, our stronger
statement and theorem quantifies Mermin's qualitative argument:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

If a QIVPM is 0-deterministic, then for a family of commuting
observables~$\left\{ \mathbf{O}_{j}\right\} _{j=1}^{M}$, we have:
\begin{equation}
\prod_{j=1}^{M}\expval{\mathbf{O}_{j}}_{\bar{\mu}}=\expval{\prod_{j=1}^{M}\mathbf{O}_{j}}_{\bar{\mu}}\label{eq:Kochen-Specker-Rules}
\end{equation}
Intuitively the expectation value of an observable lies between two
possible outcomes, which themselves lie between the minimum and
maximum bounds of the given probability intervals.  If the QIVPM is
0-deterministic, this interval collapses to a singleton
$\left\{ \lambda\right\} $, where $\lambda$ is an eigenvalue
of~$\mathbf{O}$.

@@

If we define the multiplication of two singleton
sets~$\left\{ \lambda_{1}\right\} $ and $\left\{ \lambda_{2}\right\} $
to be $\left\{ \lambda_{1}\lambda_{2}\right\} $.
